\documentclass[11pt,a4paper]{article}
\usepackage[margin=2.5cm]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{array}
\usepackage{multirow}
\usepackage{caption}

% Code listing style
\lstset{
  language=Matlab,
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{blue},
  commentstyle=\color{gray!70},
  stringstyle=\color{orange!80!black},
  frame=single,
  breaklines=true,
  numbers=left,
  numberstyle=\tiny\color{gray},
  xleftmargin=1.5em,
  framexleftmargin=1.5em,
}

\hypersetup{colorlinks=true, linkcolor=blue, citecolor=blue, urlcolor=blue}

\title{
  \Large\textbf{MATLAB 5G Toolbox Extension for 3GPP Release 19}\\[0.4em]
  \large Refined Type I Single-Panel Codebook for 128-Port Massive MIMO\\[0.4em]
  \normalsize NR Cell Performance Simulation --- 128T128R gNB / 4T4R UE
}
\author{Thang T.Q.}
\date{March 2026}

\begin{document}
\maketitle
\tableofcontents
\newpage

%% ============================================================
\section{Executive Summary}
%% ============================================================

This report documents the upgrade of the NR cell-performance MU-MIMO simulation
from 3GPP Release~15/16 (32-port gNB) to Release~19 (128-port gNB, 4-port UE).
The work involved extending the MATLAB~5G~Toolbox---specifically two internal
library files---to support the \emph{Refined Type~I Single-Panel Codebook}
defined in TS~38.214 \S5.2.2.2.1a (Rel-19).

\medskip
\noindent\textbf{Key results achieved:}
\begin{itemize}
  \item 128-port CSI-RS resource grid construction (4 resources $\times$ 32 ports,
        Row~18 CDM8, 2 slots).
  \item Full 128-port channel estimation (NMSE~$\approx -31.5$~dB at SNR~20~dB,
        CDL-C; $-27.8$~dB for CDL-B).
  \item Rel-19 modeA PMI reporting: RI selection 1--4, per-layer SINR and CQI.
  \item Demonstrated \textbf{RI~=~4} with CDL-B channel (near-uniform singular
        values), proving the codebook supports full-rank spatial multiplexing
        up to $\min(N_\text{Tx},N_\text{Rx}) = 4$ layers.
  \item Capacity comparison: SVD upper bound~37.3~bits/s/Hz vs.\
        Mode-A codebook~14.5~bits/s/Hz at RI~=~4.
\end{itemize}

%% ============================================================
\section{Background and Motivation}
%% ============================================================

\subsection{Rel-15/16 Limitation}

The existing MATLAB~5G~Toolbox (\texttt{nrPMIReport}) implements the Type~I
Single-Panel codebook of TS~38.214 \S5.2.2.2.1 for up to 32 CSI-RS ports.
For a 128-port gNB, Rel-15/16 offers no direct codebook support.

\subsection{Rel-19 Codebook Extension}

3GPP Release~19 (TS~38.214 v19.2.0) introduces \S5.2.2.2.1a,
\emph{Refined Type~I Single-Panel Codebook}, supporting 48, 64, and 128 antenna
ports via CSI-RS resource aggregation. For 128~ports:
\begin{align}
  N_1 \times N_2 \times 2 \;\text{pol} &= 128, \quad
  (N_1,N_2) \in \{(16,4),\;(8,8)\}, \quad O_1 = O_2 = 4.
\end{align}
The precoder is based on a 2D over-sampled DFT grid (modeA/modeB).

%% ============================================================
\section{Implementation Approach}
%% ============================================================

\subsection{Design Principle: Workspace Custom Library}

Rather than modifying the installed MATLAB toolbox, copies of the two relevant
internal files were placed in the project workspace:
\begin{verbatim}
  CSIRS_128T128R_v1_0/
    5g/5g/nrCSIReportConfig.m                 (copy)
    5g/5g/+nr5g/+internal/nrPMIReport.m       (copy)
    CSIRS_128T128R_v1_0.m                     (main simulation script)
\end{verbatim}
MATLAB's path resolution picks the workspace copies first, leaving the
installed toolbox untouched. All Rel-19 additions are guarded with the
comment tag \texttt{\% ThangTQ23\_128T128R\_Rel19} for traceability.

A feature flag in the main script controls the execution path:
\begin{lstlisting}
ThangTQ23_128T128R_Rel19 = true;  % true -> Approach C (Rel-19)
\end{lstlisting}

%% ============================================================
\section{File Modifications}
%% ============================================================

\subsection{\texttt{nrCSIReportConfig.m} --- Configuration Table Extension}

\textbf{Change~1: New codebook type token.}
\begin{lstlisting}
% Original (Rel-15/16):
CodebookType_Values = {'type1SinglePanel','type1MultiPanel','type2','eType2'};

% Added (Rel-19):
CodebookType_Values = {'type1SinglePanel','type1MultiPanel','type2','eType2',...
                       'typeI-SinglePanel-r19'};
\end{lstlisting}

\textbf{Change~2: 128-port entries in \texttt{getSinglePanelConfigurations()}.}
The function returns a lookup table \texttt{\{Pcsirs, [N1 N2], [O1 O2]\}}.
Six new rows were appended matching Table~5.2.2.2.1a-1:

\begin{table}[h]
\centering
\caption{New rows added to \texttt{getSinglePanelConfigurations()}}
\begin{tabular}{rcc}
\toprule
$P_\text{CSIRS}$ & $(N_1,N_2)$ & $(O_1,O_2)$ \\
\midrule
48 & (8,3) & (4,4) \\
48 & (6,4) & (4,4) \\
48 & (16,2) & (4,4) \\
64 & (8,4) & (4,4) \\
\textbf{128} & \textbf{(16,4)} & \textbf{(4,4)} \\
\textbf{128} & \textbf{(8,8)} & \textbf{(4,4)} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{\texttt{nrPMIReport.m} --- Codebook Engine Extension}

This is the core modification. Four hooks were added to existing code, and
one new function was appended (155~lines).

\subsubsection{Hook~1: Port Count Override (line~375)}

The original code reads the port count from the CSI-RS config object, which
reports 32 (one resource). For the aggregated 128-port case, the actual count
must come from the channel matrix dimension:

\begin{lstlisting}
% ThangTQ23_128T128R_Rel19: H is aggregated 128-port
if strcmpi(reportConfig.CodebookType,'typeI-SinglePanel-r19')
    numCSIRSPorts = size(H,4);   % override: 128
end
\end{lstlisting}

\subsubsection{Hook~2: CodebookType Flag Alias (line~386)}

The existing flow uses \texttt{isType1SinglePanel} to gate output formatting.
The r19 type is aliased so it inherits the same path:

\begin{lstlisting}
isType1SinglePanel_r19 = strcmpi(reportConfig.CodebookType,...
                                 'typeI-SinglePanel-r19');
if isType1SinglePanel_r19; isType1SinglePanel = true; end
\end{lstlisting}

\subsubsection{Hook~3: Codebook Dispatch in \texttt{getCodebook()} (line~476)}

\begin{lstlisting}
if strcmpi(reportConfig.CodebookType,'typeI-SinglePanel-r19')
    codebook = getPMIType1SinglePanelR19Codebook(reportConfig,...
                                                 numCSIRSPorts, nLayers);
    [~,~,i2L,i11L,i12L,i13L] = size(codebook);
    indexSetSizes = [i2L, i11L, i12L, i13L];
    return;
end
\end{lstlisting}

\subsubsection{Hook~4: PMI Wideband Selection (line~1537)}

Added r19 to the \texttt{Type1SinglePanel} branch of
\texttt{getTypeIPMIWideband()} so the 4-index \texttt{ind2sub} path
\texttt{[i2,i11,i12,i13]} is used:

\begin{lstlisting}
if strcmpi(reportConfig.CodebookType,'Type1SinglePanel') || ...
        strcmpi(reportConfig.CodebookType,'typeI-SinglePanel-r19')
    [i2,i11,i12,i13] = ind2sub(size(totalSINR),...
        find(totalSINR == max(totalSINR,[],'all'),1));
    ...
end
\end{lstlisting}

\subsubsection{New Function: \texttt{getPMIType1SinglePanelR19Codebook()} (line~2208)}

This function implements TS~38.214 Table~5.2.2.2.1a-4 (modeA) for ranks 1--4.
The codebook tensor has dimensions:
\begin{equation}
  \mathbf{C} \in \mathbb{C}^{P \times r \times N_{i_2} \times N_{i_{1,1}} \times N_{i_{1,2}} \times N_{i_{1,3}}}
\end{equation}
where $P=128$, $r$ is the rank, and the index sizes are given in
Table~5.2.2.2.1a-2.

\paragraph{Beam vector $v_{l,m}$}
\begin{equation}
  v_{l,m} = \mathbf{u}_h(l) \otimes \mathbf{u}_v(m), \quad
  [\mathbf{u}_h(l)]_k = e^{j2\pi l k / (O_1 N_1)}, \quad
  [\mathbf{u}_v(m)]_k = e^{j2\pi m k / (O_2 N_2)}
\end{equation}

\paragraph{Rank-1 precoder ($n = i_2 \in \{0,1,2,3\}$)}
\begin{equation}
  W^{(1)} = \frac{1}{\sqrt{P}}
  \begin{bmatrix} v_{l,m} \\ \varphi(n)\,v_{l,m} \end{bmatrix}
\end{equation}

\paragraph{Rank-2 precoder ($n = i_2 \in \{0,1\}$, offset $(k_1,k_2)$ from $i_{1,3}$)}
\begin{equation}
  W^{(2)} = \frac{1}{\sqrt{2P}}
  \begin{bmatrix}
    v_{l,m} & v_{l',m'} \\
    \varphi(n)\,v_{l,m} & -\varphi(n)\,v_{l',m'}
  \end{bmatrix}, \quad
  l' = l+k_1 \pmod{N_1 O_1}, \; m' = m+k_2 \pmod{N_2 O_2}
\end{equation}

\paragraph{Rank-4 precoder}
\begin{equation}
  W^{(4)} = \frac{1}{\sqrt{4P}}
  \begin{bmatrix}
    v_0 & v_1 & v_2 & v_3 \\
    \varphi(n)\,v_0 & \varphi(n)\,v_1 & \varphi(n)\,v_2 & \varphi(n)\,v_3
  \end{bmatrix}
\end{equation}
where $v_k = v_{l+k k_1,\;m+k k_2}$ for $k=0,1,2,3$.

\paragraph{Beam offsets (Table~5.2.2.2.1a-3, $N_2 > 1$)}
\begin{equation}
  i_{1,3} \in \{0,1,2,3\} \;\Rightarrow\;
  (k_1,k_2) \in \{(O_1,0),\;(0,O_2),\;(O_1,O_2),\;(0,2O_2)\}
\end{equation}

\paragraph{Code implementation of beam offsets}
\begin{lstlisting}
k1_vec = [O1  0   O1  0  ];   % i1,3 = 0,1,2,3
k2_vec = [0   O2  O2  2*O2];  % for N2 > 1
\end{lstlisting}

%% ============================================================
\section{Main Script Changes (\texttt{CSIRS\_128T128R\_v1\_0.m})}
%% ============================================================

\subsection{128-Port CSI-RS Resource Aggregation}

Four NZP-CSI-RS resources (Row~18, CDM8, 32~ports each) are configured
across 2 slots to cover all 128 ports:

\begin{table}[h]
\centering
\caption{CSI-RS resource mapping for 128-port aggregation}
\begin{tabular}{cccc}
\toprule
Resource & Slot & Symbols & Ports \\
\midrule
0 & 0 & 2,3,4,5 & 0--31 \\
1 & 0 & 8,9,10,11 & 32--63 \\
2 & 1 & 2,3,4,5 & 64--95 \\
3 & 1 & 8,9,10,11 & 96--127 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Wideband Channel Matrix Construction}

Each per-resource channel estimate
$\hat{\mathbf{H}}_r \in \mathbb{C}^{K \times L \times 4 \times 32}$
is averaged over frequency and time and assembled into:
\begin{equation}
  \mathbf{H}_\text{wb} \in \mathbb{C}^{4 \times 128}, \quad
  [\mathbf{H}_\text{wb}]_{:,\,pS:pE} =
    \mathbb{E}_{k,l}\left[\hat{\mathbf{H}}_r(k,l,:,:)\right]
\end{equation}
This is broadcast to $[K \times L \times 4 \times 128]$ for \texttt{nrPMIReport}.

\subsection{RI Selection via Capacity Metric}

\textbf{Critical design choice:} RI is selected by maximising total achievable rate
(log-domain), \emph{not} linear SINR sum. Using linear SINR sum would always
favour RI~=~1 because a single high-SINR layer dominates the sum:
\begin{lstlisting}
for ri_try = 1:min(nRxAntennas, 4)
    [~, info_try] = nr5g.internal.nrPMIReport(..., ri_try, ...);
    sinr_vals = info_try.SINRPerREPMI(:);
    rate_sum  = sum(log2(1 + sinr_vals(isfinite(sinr_vals)...
                                       & sinr_vals >= 0)));
    if rate_sum > best_rate
        best_rate = rate_sum;
        best_ri   = ri_try;
    end
end
\end{lstlisting}

%% ============================================================
\section{Verification Against TS~38.214 Rel-19}
%% ============================================================

\begin{table}[h]
\centering
\caption{Spec compliance verification}
\begin{tabular}{lp{6cm}l}
\toprule
Spec requirement & Code reference & Status \\
\midrule
Table 5.2.2.2.1a-1: $(N_1,N_2) \in \{(16,4),(8,8)\}$, $O_1=O_2=4$ for 128p
  & \texttt{nrCSIReportConfig.m}, config table & \checkmark \\
Table 5.2.2.2.1a-2: $i_{1,1} \in [0..63]$, $i_{1,2} \in [0..15]$ (modeA, $(16,4)$)
  & \texttt{i11\_len=N1*O1}, \texttt{i12\_len=N2*O2} & \checkmark \\
Table 5.2.2.2.1a-3: $(k_1,k_2)$ offsets for rank 2--4
  & \texttt{k1\_vec, k2\_vec} arrays & \checkmark \\
Table 5.2.2.2.1a-4: Rank-1 precoder formula
  & \texttt{[vlm; phi(i2)*vlm]/sqrt(P)} & \checkmark \\
Table 5.2.2.2.1a-4: Rank-2 precoder formula
  & \texttt{[vlm,vp; pn*vlm,-pn*vp]/sqrt(2P)} & \checkmark \\
Table 5.2.2.2.1a-4: Rank-4 precoder formula
  & \texttt{[v0..v3; pn*v0..pn*v3]/sqrt(4P)} & \checkmark \\
$\varphi(n) = e^{j\pi n/2}$
  & \texttt{phi = @(x) exp(1i*pi*x/2)} & \checkmark \\
$v_{l,m} = \mathbf{u}_h(l) \otimes \mathbf{u}_v(m)$ (DFT)
  & \texttt{getVlm()}: \texttt{reshape((ul.*um).',[],1)} & \checkmark \\
Normalisation $1/\sqrt{r P}$
  & \texttt{1/sqrt(r*Pcsirs)} for $r=1,2,3,4$ & \checkmark \\
\bottomrule
\end{tabular}
\end{table}

\noindent\textbf{Known limitations (future work):}
\begin{itemize}
  \item Ranks 5--8: not implemented; code falls back to rank-4 with warning.
  \item modeB: not implemented (requires independent per-layer beam search).
\end{itemize}

%% ============================================================
\section{Simulation Results}
%% ============================================================

\subsection{Simulation Parameters}

\begin{table}[h]
\centering
\caption{Simulation configuration}
\begin{tabular}{ll}
\toprule
Parameter & Value \\
\midrule
Carrier & 52 PRBs, SCS 30~kHz, 3.5~GHz \\
gNB antenna & $4V \times 16H \times 2\text{pol} = 128$~ports \\
UE antenna & $1V \times 2H \times 2\text{pol} = 4$~ports \\
CSI-RS & 4 resources, Row~18, CDM8, 32~ports each \\
Channel (rank-4 test) & CDL-B, DS~100~ns, $f_D = 5$~Hz \\
Channel (original) & CDL-C, DS~100~ns, $f_D = 5$~Hz \\
SNR & 20~dB \\
Codebook & typeI-SinglePanel-r19 (modeA) \\
$(N_1,N_2)$ & $(16,4)$, $O_1=O_2=4$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Channel Estimation Quality}

\begin{table}[h]
\centering
\caption{Per-resource channel estimation NMSE}
\begin{tabular}{ccc}
\toprule
Channel & Overall MSE & NMSE \\
\midrule
CDL-C & $4.90 \times 10^{-4}$ & $-31.5$~dB \\
CDL-B & $6.64 \times 10^{-4}$ & $-27.8$~dB \\
\bottomrule
\end{tabular}
\end{table}

\subsection{PMI Reporting Results}

\begin{table}[h]
\centering
\caption{Rel-19 modeA PMI report --- CDL-B channel, SNR~20~dB}
\begin{tabular}{lc}
\toprule
Metric & Value \\
\midrule
RI selected & \textbf{4} \\
PMI $i_1$ (0-based) & $[57, 14, 0]$ \\
PMI $i_2$ (0-based) & 1 \\
Per-layer SINR & $[12.8,\;13.7,\;14.6,\;16.8]$~dB \\
Per-layer CQI (64-QAM mapping) & $[10,\;10,\;11,\;12]$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Capacity Comparison}

\begin{table}[h]
\centering
\caption{Capacity at wideband channel (CDL-B, SNR~20~dB)}
\begin{tabular}{lccc}
\toprule
Approach & RI & Capacity (bits/s/Hz) & Gap to SVD \\
\midrule
SVD (upper bound) & 4 & 37.31 & --- \\
Rel-19 modeA codebook & 4 & 14.51 & $-22.8$~dB \\
\bottomrule
\end{tabular}
\end{table}

The gap between SVD and modeA (22.8~bits/s/Hz) is expected: the DFT codebook
constrains the second, third, and fourth beams to fixed offsets $(k_1,k_2)$
relative to the dominant beam, whereas SVD uses the exact right singular
vectors of $\mathbf{H}$.

\subsection{SVD Channel Rank Analysis (CDL-B)}

\begin{table}[h]
\centering
\caption{Singular value decomposition of wideband $\mathbf{H}$ (CDL-B)}
\begin{tabular}{cccc}
\toprule
SV \# & Singular value & SNR (dB) & Capacity contribution (bits/s/Hz) \\
\midrule
1 & 8.37 & 32.1 & 10.66 \\
2 & 5.94 & 29.1 & 9.67 \\
3 & 4.18 & 26.0 & 8.65 \\
4 & 3.75 & 25.1 & 8.34 \\
\midrule
\multicolumn{3}{l}{\textbf{Total SVD capacity}} & \textbf{37.31} \\
\bottomrule
\end{tabular}
\end{table}

CDL-B exhibits near-uniform singular values (only 7~dB spread across 4 SVs),
confirming that the channel supports spatial multiplexing up to rank~4.

%% ============================================================
\section{Conclusions}
%% ============================================================

The simulation framework has been successfully upgraded from Rel-15/16 to
Release~19 with the following verified outcomes:

\begin{enumerate}
  \item \textbf{128-port CSI-RS}: Resource aggregation (4$\times$32~ports,
        Row~18 CDM8) correctly generates 128-port channel estimates at
        NMSE~$< -27$~dB.

  \item \textbf{Rel-19 Codebook}: The Refined Type~I Single-Panel codebook
        (TS~38.214 \S5.2.2.2.1a, modeA) has been implemented in the MATLAB
        workspace library with full spec compliance for ranks~1--4.

  \item \textbf{RI~=~4 Demonstrated}: With CDL-B channel (near-uniform SVs),
        the Rel-19 modeA codebook correctly selects RI~=~4 and reports
        per-layer SINR $\approx 12$--17~dB at SNR~20~dB.

  \item \textbf{No Toolbox Corruption}: All modifications are isolated to
        workspace copies; the installed MATLAB 5G Toolbox is unchanged.
\end{enumerate}

\noindent\textbf{Next steps:}
\begin{itemize}
  \item Implement modeB (independent beam search per layer) for improved
        multi-layer performance on rank-sparse channels (e.g.\ CDL-C).
  \item Compare with Huawei R1-2500098 Scheme-B (greedy orthogonal DFT)
        which is the modeB reference for the Rel-19 study item.
  \item Extend to multi-user (MU-MIMO) scheduling with the 128-port codebook.
\end{itemize}

\end{document}
